---
title: "Microbiome OCP - Alpha Diversity"
author: "Fernanda Terrazas"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)

```
# 2. Alpha Diversity
## 2.1 log-transform alpha diversity metrics
```{r}
alpha_diversity_tx <- alpha_diversity
head(alpha_diversity_tx)
alpha_diversity_tx[, 2:5] <- sqrt(alpha_diversity_tx[, 2:5])
head(alpha_diversity_tx)

shapiro.test(alpha_diversity_tx$shannon_entropy)
shapiro.test(alpha_diversity_tx$observed_features)
shapiro.test(alpha_diversity_tx$faith_pd)
shapiro.test(alpha_diversity_tx$pielou_evenness)

```
Assign data subset to "group" variable to use for downstream analyses
```{r}
group <- group_control_oc
group

```
### Add metadata to alpha diversity data
```{r}
# Merge alpha diversity data with assigned group
alpha_merged <- merge(alpha_diversity_tx, group, by = "SampleID")

```
### Linear Model -- Effect of **Treatment * Day** on shannon_entropy, observed_features, faith_pd", pielou_evenness
```{r}
library(MuMIn)

# Fit model for each alpha diversity metric, where Treatment is the group variable and 1|SubjectID specifies the random effect
lmer_shannon <- lmer(alpha_merged$shannon_entropy ~ Treatment * Day + (1|Subject), data = alpha_merged)
summary(lmer_shannon)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_shannon <- lmer(alpha_merged$shannon_entropy ~ Treatment * Day + (1|Subject), data = alpha_merged)
lmer_shannon_anova <- Anova(lmer_shannon, type = "III")
lmer_shannon_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_shannon)

lmer_obsft <- lmer(alpha_merged$observed_features ~ Treatment * Day + (1|Subject), data = alpha_merged)
summary(lmer_obsft)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit
lmer_obsft <- lmer(alpha_merged$observed_features ~ Treatment * Day + (1|Subject), data = alpha_merged)
lmer_obsft_anova <- Anova(lmer_obsft, type = "III")
lmer_obsft_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_obsft)


lmer_faith <- lmer(alpha_merged$faith_pd ~ Treatment * Day + (1|Subject), data = alpha_merged)
summary(lmer_faith)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_faith <- lmer(alpha_merged$faith_pd ~ Treatment * Day + (1|Subject), data = alpha_merged)
lmer_faith_anova <- Anova(lmer_faith, type = "III")
lmer_faith_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_faith)


lmer_pielou <- lmer(alpha_merged$pielou_evenness ~ Treatment * Day + (1|Subject), data = alpha_merged)
summary(lmer_pielou)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_pielou <- lmer(alpha_merged$pielou_evenness ~ Treatment * Day + (1|Subject), data = alpha_merged)
lmer_pielou_anova <- Anova(lmer_pielou, type = "III")
lmer_pielou_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_pielou)

```
Combine lme results for all alpha diversity metrics in a single df to save
```{r}
lmer_shannon_df <- lmer_shannon_anova
lmer_obsft_df <- lmer_obsft_anova
lmer_faith_df <- lmer_faith_anova
lmer_pielou_df <- lmer_pielou_anova

```
Save lme model results as .csv file
```{r}
write.csv(lmer_shannon_df, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-results_lmer_shannon_df.csv", row.names = FALSE)

write.csv(lmer_obsft_df, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-results_lmer_obsft_df.csv", row.names = FALSE)

write.csv(lmer_faith_df, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-results_lmer_faith_df.csv", row.names = FALSE)

write.csv(lmer_pielou_df, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-results_lmer_pielou_df.csv", row.names = FALSE)

```
### Check assumptions of the model **Treatment & Day**

#### 1. Linearity - the relationship between the dependent variable and the fixed effects should be linear.
                  - the relationship between predictors and the dependent variable is linear.
```{r}
par(mfrow = c(2, 2))

plot(resid(lmer_shannon), alpha_merged$shannon_entropy)
plot(resid(lmer_obsft), alpha_merged$observed_features)
plot(resid(lmer_faith), alpha_merged$faith_pd)
plot(resid(lmer_pielou), alpha_merged$pielou_evenness)

```
#### 2. Homoscedasticity - the variance of residuals should be constant across the range of fitted values.
                         - the residuals have constant variance.
```{r}
# Plot residuals vs. fitted values
plot(resid(lmer_shannon) ~ fitted(lmer_shannon)) # or plot(lmer_shannon)
plot(resid(lmer_obsft) ~ fitted(lmer_obsft))
plot(resid(lmer_faith) ~ fitted(lmer_faith))
plot(resid(lmer_pielou) ~ fitted(lmer_pielou))

```
#### Normality - the residuals of the model are normally distributed
```{r}
# id= identifies values that may be exerting undue influence on the model (i.e. outliers)
qqmath(lmer_shannon, id=0.05)
qqmath(lmer_obsft, id=0.05)
qqmath(lmer_faith, id=0.05)
qqmath(lmer_pielou, id=0.05)

```
#### 5. No multicollinearity - the predictor variables are not perfectly correlated with each other.

A VIF greater than 5 or 10 often indicates problematic multicollinearity. In mixed models, you can calculate VIF for the fixed effects only, because the random effects are assumed to be uncorrelated with the fixed effects and with each other:
```{r}
library(car)

vif(lmer_shannon)
vif(lmer_obsft)
vif(lmer_faith)
vif(lmer_pielou)

```

```{r}
# Reset the mfrow parameter to default
par(mfrow = c(1, 1))

```
### Linear Model -- Effect of **Diet** on shannon_entropy, observed_features, faith_pd", pielou_evenness
Subset alpha_merged according to Prune-NoPrune group
```{r}
group <- group_prune_no.prune

# Merge alpha diversity data with assigned group
alpha_merged_prune <- merge(alpha_diversity_tx, group, by = "SampleID")

```


```{r}
# Fit model for each alpha diversity metric, where Diet is the group variable and 1|SubjectID specifies the random effect
lmer_shannon_diet <- lmer(alpha_merged_prune$shannon_entropy ~ Diet + (1|Subject), data = alpha_merged_prune)
summary(lmer_shannon_diet)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_shannon_diet <- lmer(alpha_merged_prune$shannon_entropy ~ Diet + (1|Subject), data = alpha_merged_prune)
lmer_shannon_diet_anova <- Anova(lmer_shannon_diet, type = "III")
lmer_shannon_diet_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_shannon_diet)


lmer_obsft_diet <- lmer(alpha_merged_prune$observed_features ~ Diet + (1|Subject), data = alpha_merged_prune)
summary(lmer_obsft_diet)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_obsft_diet <- lmer(alpha_merged_prune$observed_features ~ Diet + (1|Subject), data = alpha_merged_prune)
lmer_obsft_diet_anova <- Anova(lmer_obsft_diet, type = "III")
lmer_obsft_diet_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_obsft_diet)


lmer_faith_diet <- lmer(alpha_merged_prune$faith_pd ~ Diet + (1|Subject), data = alpha_merged_prune)
summary(lmer_faith_diet)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_faith_diet <- lmer(alpha_merged_prune$faith_pd ~ Diet + (1|Subject), data = alpha_merged_prune)
lmer_faith_diet_anova <- Anova(lmer_faith_diet, type = "III")
lmer_faith_diet_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_faith_diet)


lmer_pielou_diet <- lmer(alpha_merged_prune$pielou_evenness ~ Diet + (1|Subject), data = alpha_merged_prune)
summary(lmer_pielou_diet)
options(contrasts = c("contr.sum","contr.poly")) #change contrasts and refit model
lmer_pielou_diet <- lmer(alpha_merged_prune$pielou_evenness ~ Diet + (1|Subject), data = alpha_merged_prune)
lmer_pielou_diet_anova <- Anova(lmer_pielou_diet, type = "III")
lmer_pielou_diet_anova
options(contrasts = c("contr.treatment","contr.poly")) #change contrasts back
r.squaredGLMM(lmer_pielou_diet)

```
### Check assumptions of the model **Diet**

#### 1. Linearity - the relationship between the dependent variable and the fixed effects should be linear.
                  - the relationship between predictors and the dependent variable is linear.
```{r}
par(mfrow = c(2, 2))

plot(resid(lmer_shannon_diet), alpha_merged_prune$shannon_entropy)
plot(resid(lmer_obsft_diet), alpha_merged_prune$observed_features)
plot(resid(lmer_faith_diet), alpha_merged_prune$faith_pd)
plot(resid(lmer_pielou_diet), alpha_merged_prune$pielou_evenness)

```
#### 2. Homoscedasticity - the variance of residuals should be constant across the range of fitted values.
                         - the residuals have constant variance.
```{r}
# Plot residuals vs. fitted values
plot(resid(lmer_shannon_diet) ~ fitted(lmer_shannon_diet)) # or plot(lmer_shannon_diet)
plot(resid(lmer_obsft_diet) ~ fitted(lmer_obsft_diet))
plot(resid(lmer_faith_diet) ~ fitted(lmer_faith_diet))
plot(resid(lmer_pielou_diet) ~ fitted(lmer_pielou_diet))

```
#### Normality - the residuals of the model are normally distributed
```{r}
# id= identifies values that may be exerting undue influence on the model (i.e. outliers)
qqmath(lmer_shannon_diet, id=0.05)
qqmath(lmer_obsft_diet, id=0.05)
qqmath(lmer_faith_diet, id=0.05)
qqmath(lmer_pielou_diet, id=0.05)

```
#### 5. No multicollinearity - the predictor variables are not perfectly correlated with each other.

**N/A since only testing Diet effects here**

```{r}
# Reset the mfrow parameter to default
par(mfrow = c(1, 1))

```
### Plot alpha diversity
```{r}
# merge alpha diversity data with all data
alpha_all <- merge(alpha_diversity_tx, map, by = "SampleID")

# Create a data frame for plotting
alpha_plot_data <- data.frame(
  Metric = rep(c("Shannon Entropy", "Observed Features", "Faith PD", "Pielou Evenness"), each = nrow(map)),
  Value = c(alpha_all$shannon_entropy, alpha_all$observed_features, alpha_all$faith_pd, alpha_all$pielou_evenness),
  Treatment = rep(alpha_all$Treatment, times = 4),
  Diet = rep(alpha_all$Diet, times = 4),
  Day = rep(alpha_all$Day, times = 4)
)
head(alpha_plot_data)

```

```{r}
# append Day to every entry in the 'Day' column for better label aesthetics
# alpha_plot_data <- alpha_plot_data %>%
#  mutate(Day = paste0("Day ", Day))

# Create box plots using 'Day' as a facet
# alpha_treat_p <- ggplot(alpha_plot_data, aes(x = Treatment, y = Value, fill = Treatment)) +
#   geom_boxplot(colour = "black") +
#   scale_fill_manual(values = c( "#332288","#E066A6", "#E6A316", "#256769")) +
#   facet_grid(Metric ~ Day, scales = "free_y") +
#   theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5), # strip.text.x or strip.text.y to change x- and y-axis labels
#         axis.text.x = element_text(size = 7), # Increase x-axis tick label text size
#         axis.text.y = element_text(size = 7)) +
#   xlab("") + ylab("") +
#   ggtitle("")
# alpha_treat_p

alpha_treat_p <- ggplot(alpha_plot_data, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(colour = "black") +
  scale_fill_manual(values = c("#E6A316", "#E066A6")) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 4, strip.position = "right") +
  theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5), # strip.text.x or strip.text.y to change x- and y-axis labels
        axis.text.x = element_text(size = 7), # Increase x-axis tick label text size
        axis.text.y = element_text(size = 7)) +
  xlab("") + ylab("") +
  ggtitle("")
alpha_treat_p

alpha_treat_p <- ggplot(alpha_plot_data, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(aes(position = Day), colour = "black", position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("#E6A316", "#E066A6")) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 4, strip.position = "right") +
  theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5), # strip.text.x or strip.text.y to change x- and y-axis labels
        axis.text.x = element_text(size = 7), # Increase x-axis tick label text size
        axis.text.y = element_text(size = 7)) +
  xlab("") + ylab("") +
  ggtitle("")
alpha_treat_p

alpha_treat_p <- ggplot(alpha_plot_data, aes(x = Treatment, y = Value, fill = interaction(Treatment, Day))) +
  geom_boxplot(position = "dodge", colour = "black") +
  scale_fill_manual(values = c("#E6A316", "#E066A6", "#B6A316", "#B066A6")) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 4, strip.position = "right") +
  theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7)) +
  xlab("") + ylab("") +
  ggtitle("")
alpha_treat_p

alpha_treat_p <- ggplot(alpha_plot_data, aes(x = Treatment, y = Value, fill = interaction(Treatment, Day))) +
  geom_boxplot(position = "dodge", colour = "black") +
  scale_fill_manual(
    values = c("#E6A316", "#E066A6", "#e8d1a0", "#e0bccf"), 
    name = "Treatment:Day",
    labels = c("Control - 1", "User - 1", "Control - 21", "User - 21") # Modify as needed
  ) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 4, strip.position = "right") +
  theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7)) +
  xlab("") + ylab("") +
  ggtitle("")
alpha_treat_p



# alpha_diet_p <- ggplot(alpha_plot_data, aes(x = Diet, y = Value, fill = Diet)) +
#   geom_boxplot(colour = "black") +
#   scale_fill_manual(values = c("#E066A6","#332288", "#E6A316", "#256769")) +
#   facet_wrap(~ Metric, scales = "free_y", nrow = 4, strip.position = "right") +
#   theme(aspect.ratio = 2.5, strip.text = element_text(size = 7.5), # strip.text.x or strip.text.y to change x- and y-axis labels
#         axis.text.x = element_text(size = 7), # Increase x-axis tick label text size
#         axis.text.y = element_text(size = 7)) +
#   xlab("") + ylab("")
alpha_diet_p

# install.packages("grid")
# install.packages("egg")
# library(grid)
# library(egg)

# alpha_p_title <- grid::textGrob("", gp = gpar(fontsize = 20)) # adds a title to whole figure
# 
# alpha_treat_diet_p <- egg::ggarrange(alpha_treat_p, alpha_diet_p, labels = c("A", "B"), 
#                          ncol = 2, nrow = 1, heights = c(3), 
#                          top = alpha_p_title) # top = alpha_p_title uses the title in the figure

alpha_treat_diet_p <- egg::ggarrange(alpha_treat_p, alpha_diet_p, labels = c("A", "B"), 
                         ncol = 2, nrow = 1, heights = c(2))
alpha_treat_diet_p

```
Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-treatment-plot.png", plot = alpha_treat_p, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-diet-plot.png", plot = alpha_diet_p, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/alpha/alpha-treat_diet-plot.png", plot = alpha_treat_diet_p, width = 7, height = 5, dpi = 300)

```
