---
title: "coda4microbiome -- NO OUTLIERS"
author: "Fernanda Terrazas"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
Set working directory
```{r}
# setwd("~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses")

```
Load coda4microbiome package
```{r Load Packages}
library(coda4microbiome)

```

```{r}
# to reproduce the results
set.seed(123)

```
# 4. coda4microbiome - Treatment and Diet (binary outcome)
## 4.1 Control vs. OC Users
Assign "group" with metadata subset

**EXCLUDE H2 & CH4 OUTLIERS FROM MODELS**
```{r group}
remove_outliers <- c("950N", "950N2", "962P", "962P2", "921C", "921C2", "735P", "735P2") #  "989N", "737P"
group <- group_control_oc[!(group_control_oc$SampleID %in% remove_outliers),]

```

```{r}
# transpose asv/species/genus table -- use species for original names, species_c4m for simplified names
# x_species_cu_o <- data.frame(t(species_c4m), check.names = T)
x_species_cu_o <- data.frame(t(species), check.names = T)

# subset asv/species/genus data to keep subgroup data
x_species_cu_o <- x_species_cu_o[row.names(x_species_cu_o) %in% group$SampleID, ]

# Use group metadata subset to save Treatment column as a factor
y_treatment_cu_o <- as.factor(group$Treatment)
y_treatment_cu_o

```

```{r}
# identification of microbial signatures between Treatments (C and N, or N and P)
coda_glmnet_cu_o <- coda_glmnet(x=x_species_cu_o, y=y_treatment_cu_o, nfolds = 6, lambda = "lambda.min")
coda_glmnet_cu_o$`apparent AUC`
coda_glmnet_cu_o

# permutational test? 
# coda_glmnet_perm_cu_o <- coda_glmnet_null(x=x_species_cu_o, y=y_treatment_cu_o, niter = 100)
# coda_glmnet_perm_cu_o

```
Extract glmnet results and add to new data frame
```{r}
keep_elem <- c(1, 2, 3, 5, 6, 7) # keep elements containing taxa numbers, names, log and AUC values, etc
coda_list_cu_o <- list(coda_glmnet_cu_o[keep_elem])
coda_df_cu_o <- do.call(cbind.data.frame, coda_list_cu_o)
View(coda_df_cu_o)

```
Save coda_glmnet results as .csv file
```{r}
write.csv(coda_df_cu_o, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/coda-glmnet_results_cu_o.csv", row.names = FALSE)

```
Plot selected taxa and their estimated regression coefficients manually 
```{r}
# library(cowplot)

# Create a new data frame to create the plot
coeff_df_cu_o <- data.frame(
  taxa = coda_glmnet_cu_o$taxa.name,
  # short_taxa = short_taxa_meth,
  coeff = coda_glmnet_cu_o$`log-contrast coefficients`
)

####
#### simplify names
####

# # Assuming your data is in a data frame called 'data'
# coeff_df_cu_o <- data.frame(coeff_df_cu_o %>%
#   group_by(Species) %>%
#   summarise(coeff = sum(coeff)))

# Add a column indicating whether log_contrast is positive or negative
coeff_df_cu_o$sign <- ifelse(coeff_df_cu_o$coeff > 0, "positive", "negative")

coeff_df_cu_o_cutoff <- coeff_df_cu_o[coeff_df_cu_o$coeff > 0.05 | coeff_df_cu_o$coeff < -0.05,]

# Find the maximum absolute coefficient value
max_abs_coeff_cu_o <- max(abs(c(min(coeff_df_cu_o_cutoff$coeff), max(coeff_df_cu_o_cutoff$coeff))))

# Create the plot with short_taxa on the y-axis, log_contrast on the x-axis, and fill based on sign
p_coda_glmnet_cu_o <- ggplot(coeff_df_cu_o_cutoff, aes(x = coeff, y = reorder(Species, coeff), fill = sign)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#E6A316","#E066A6","#332288","#256769")) +
  scale_x_continuous(limits = c(-max_abs_coeff_cu_o, max_abs_coeff_cu_o)) +
  #geom_text(aes(label = round(coeff, 2)), hjust = -0.1, size = 3) +
  #theme(axis.text.y = element_text(hjust = 1)) +
  labs(x = "Log-contrast coefficient", y = "Taxa", title = "") +
  theme_classic() +
  theme(legend.position = "none")
p_coda_glmnet_cu_o

```
Create predictions plot
```{r}
# Prepare the data
predictions_cu_o <- data.frame(Value = coda_glmnet_cu_o$predictions)
predictions_cu_o$Group <- ifelse(predictions_cu_o$Value > 0, 'User', 'Control')

# Find the maximum absolute value
max_abs_value_cu_o_pred <- max(abs(c(min(predictions_cu_o$Value[predictions_cu_o$Group == "Control"]), max(predictions_cu_o$Value[predictions_cu_o$Group == "User"]))))

# # Plot the data and set the y-axis limits to be symmetrical around zero
p_predictions_cu_o <- ggplot(predictions_cu_o, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(colour = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#E6A316","#E066A6","#332288","#256769")) +
  scale_y_continuous(limits = c(-max_abs_value_cu_o_pred, max_abs_value_cu_o_pred)) +
  theme_minimal() +
  labs(title = " ",
       x = "",
       y = "Balance value") +
  theme_classic() +
  theme(legend.position = "none")
p_predictions_cu_o

```
Combine 
```{r}
p_coda_glmnet_plots_cu_o <- egg::ggarrange(p_coda_glmnet_cu_o, p_predictions_cu_o, labels = c("", ""), 
                         ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_glmnet_plots_cu_o)
 
```
Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/signature-plot_cu_o.png", plot = p_coda_glmnet_cu_o, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_cu_o.png", plot = p_coda_glmnet_plots_cu_o, width = 6, height = 4, dpi = 300) 

```
## 4.2 Prune vs. No Prune
Assign "group" with metadata subset

**EXCLUDE H2 & CH4 OUTLIERS FROM MODELS**
```{r group}
group <- group_prune_no.prune[!(group_prune_no.prune$SampleID %in% remove_outliers),]

```

```{r}
# transpose asv/species/genus table -- use species for original names, species_c4m for simplified names
# x_species_pnp_o <- data.frame(t(species_c4m), check.names = T)
x_species_pnp_o <- data.frame(t(species), check.names = T)

# subset asv/species/genus data to keep subgroup data
x_species_pnp_o <- x_species_pnp_o[row.names(x_species_pnp_o) %in% group$SampleID, ]

# Use group metadata subset to save Treatment column as a factor
y_treatment_pnp_o <- as.factor(group$Diet)
y_treatment_pnp_o

```

```{r}
# identification of microbial signatures between Treatments (C and N, or N and P)
coda_glmnet_pnp_o <- coda_glmnet(x=x_species_pnp_o, y=y_treatment_pnp_o, nfolds = 6, lambda = "lambda.min")
coda_glmnet_pnp$`apparent AUC`

# permutational test? 
# coda_glmnet_perm_pnp_o <- coda_glmnet_null(x=x_species_pnp_o, y=y_treatment_pnp_o, niter = 100)
# coda_glmnet_perm_pnp_o

```
Extract glmnet results and add to new data frame
```{r}
keep_elem <- c(1, 2, 3, 5, 6, 7) # keep elements containing taxa numbers, names, log and AUC values, etc
coda_list_pnp_o <- list(coda_glmnet_pnp_o[keep_elem])
coda_df_pnp_o <- do.call(cbind.data.frame, coda_list_pnp_o)
View(coda_df_pnp_o)

```
Save coda_glmnet results as .csv file
```{r}
write.csv(coda_df_pnp_o, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/coda-glmnet_results_pnp_o.csv", row.names = FALSE)

```
Plot selected taxa and their estimated regression coefficients manually 
```{r}
# Create a new data frame to create the plot
coeff_df_pnp_o <- data.frame(
  taxa = coda_glmnet_pnp_o$taxa.name,
  # short_taxa = short_taxa_meth,
  coeff = coda_glmnet_pnp_o$`log-contrast coefficients`
)

####
#### simplify names
####

# # Assuming your data is in a data frame called 'data'
# coeff_df_pnp_o <- data.frame(coeff_df_pnp_o %>%
#   group_by(Species) %>%
#   summarise(coeff = sum(coeff)))

# Add a column indicating whether log_contrast is positive or negative
coeff_df_pnp_o$sign <- ifelse(coeff_df_pnp_o$coeff > 0, "positive", "negative")

coeff_df_pnp_o_cutoff <- coeff_df_pnp_o[coeff_df_pnp_o$coeff > 0.05 | coeff_df_pnp_o$coeff < -0.05,]

# Find the maximum absolute coefficient value
max_abs_coeff_pnp_o <- max(abs(c(min(coeff_df_pnp_o_cutoff$coeff), max(coeff_df_pnp_o_cutoff$coeff))))

# Create the plot with short_taxa on the y-axis, log_contrast on the x-axis, and fill based on sign
p_coda_glmnet_pnp_o <- ggplot(coeff_df_pnp_o_cutoff, aes(x = coeff, y = reorder(Species, coeff), fill = sign)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#E066A6","#332288","#E6A316","#256769")) +
  scale_x_continuous(limits = c(-max_abs_coeff_pnp_o, max_abs_coeff_pnp_o)) +
  #geom_text(aes(label = round(coeff, 2)), hjust = -0.1, size = 3) +
  #theme(axis.text.y = element_text(hjust = 1)) +
  labs(x = "Log-contrast coefficient", y = "Taxa", title = "") +
  theme_classic() +
  theme(legend.position = "none")
p_coda_glmnet_pnp_o

```
Create predictions plot
```{r}
# Prepare the data
predictions_pnp_o <- data.frame(Value = coda_glmnet_pnp_o$predictions)
predictions_pnp_o$Group <- ifelse(predictions_pnp_o$Value > 0, 'Prune', 'No Prune')

# Find the maximum absolute value
max_abs_value_pnp_o_pred <- max(abs(c(min(predictions_pnp_o$Value[predictions_pnp_o$Group == "No Prune"]), max(predictions_pnp_o$Value[predictions_pnp_o$Group == "Prune"]))))

# # Plot the data and set the y-axis limits to be symmetrical around zero
p_predictions_pnp_o <- ggplot(predictions_pnp_o, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(colour = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#E066A6","#332288","#E6A316","#256769")) +
  scale_y_continuous(limits = c(-max_abs_value_pnp_o_pred, max_abs_value_pnp_o_pred)) +
  theme_minimal() +
  labs(title = " ",
       x = "",
       y = "Balance value") +
  theme_classic() +
  theme(legend.position = "none")
p_predictions_pnp_o

```
Combine 
```{r}
p_coda_glmnet_plots_pnp_o <- egg::ggarrange(p_coda_glmnet_pnp_o, p_predictions_pnp_o, labels = c("", ""), 
                         ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_glmnet_plots_pnp_o)
 
```
Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/signature-plot_pnp_o.png", plot = p_coda_glmnet_pnp_o, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_pnp_o.png", plot = p_coda_glmnet_plots_pnp_o, width = 6, height = 4, dpi = 300) 
```
# 5. coda4microbiome - CH4 & H2 signatures in Prune (continuous variables into binary outcome)
Assign "group" with metadata subset
```{r}
group <- map[!(map$SampleID %in% remove_outliers),]

```
subset species data
```{r}
# neet to transpose asv/species/genus table for coda_glmnet()
# x_species_p <- data.frame(t(species_c4m), check.names = T)
x_species_p <- data.frame(t(species), check.names = T)

# subset asv/species/genus data to keep subgroup data
x_species_p <- x_species_p[row.names(x_species_p) %in% group$SampleID, ]

```
### Use group and complete.cases() to save CH4 and H2 columns that have data for all entries
```{r}
y_methane_p <- subset(group, complete.cases(CH4))
y_methane_p$CH4 <- sqrt(y_methane_p$CH4)
mean_CH4 <- mean(y_methane_p$CH4) # calculate the mean of CH4
y_methane_p$b_CH4 <- ifelse(y_methane_p$CH4 > mean_CH4, "high", "low") # create a b_CH4 column with high/low values for CH4
y_methane_pb <- y_methane_p$b_CH4
x_species_meth_p <- x_species_p[row.names(x_species_p) %in% y_methane_p$SampleID, ]


y_hydrogen_p <- subset(group, complete.cases(H2))
y_hydrogen_p$H2 <- sqrt(y_hydrogen_p$H2)
mean_H2 <- mean(y_hydrogen_p$H2) # calculate the mean of H2
y_hydrogen_p$b_H2 <- ifelse(y_hydrogen_p$H2 > mean_H2, "high", "low") # create a b_H2 column with high/low values for H2
y_hydrogen_pb <- y_hydrogen_p$b_H2
x_species_hydro_p <- x_species_p[row.names(x_species_p) %in% y_hydrogen_p$SampleID, ]

```

```{r}
# identification of microbial signatures
coda_glmnet_meth_p <- coda_glmnet(x=x_species_meth_p, y=y_methane_pb, nfolds = 7, lambda = "lambda.min")
coda_glmnet_meth_p$`apparent AUC`
coda_glmnet_meth_p

coda_glmnet_hydro_p <- coda_glmnet(x=x_species_hydro_p, y=y_hydrogen_pb, nfolds = 6, lambda = "lambda.min")
coda_glmnet_hydro_p$`apparent AUC`
coda_glmnet_hydro_p

```
Extract glmnet results and add to new data frame
```{r}
keep_elem <- c(1, 2, 3, 5, 6, 7) # keep elements containing taxa numbers, names, log and AUC values, etc
coda_list_meth_p <- list(coda_glmnet_meth_p[keep_elem])
coda_df_meth_p <- do.call(cbind.data.frame, coda_list_meth_p)
View(coda_df_meth_p)

keep_elem <- c(1, 2, 3, 5, 6, 7) # keep elements containing taxa numbers, names, log and AUC values, etc
coda_list_hydro_p <- list(coda_glmnet_hydro_p[keep_elem])
coda_df_hydro_p <- do.call(cbind.data.frame, coda_list_hydro_p)
View(coda_df_hydro_p)

```
Save coda_glmnet results as .csv file
```{r}
write.csv(coda_df_meth_p, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/glmnet-results_meth.csv", row.names = FALSE)

write.csv(coda_df_hydro_p, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/glmnet-results_hydro.csv", row.names = FALSE)

```
**CH4** -- Plot selected taxa and their estimated regression coefficients
```{r}
# library(cowplot)

# Create a new data frame to create the plot
coeff_df_meth_p <- data.frame(
  taxa = coda_glmnet_meth_p$taxa.name,
  # short_taxa = short_taxa_meth,
  coeff = coda_glmnet_meth_p$`log-contrast coefficients`
)

####
#### simplify names
####

# # Assuming your data is in a data frame called 'data'
# coeff_df_meth_p <- data.frame(coeff_df_meth_p %>%
#   group_by(Species) %>%
#   summarise(coeff = sum(coeff)))

# Add a column indicating whether log_contrast is positive or negative
coeff_df_meth_p$sign <- ifelse(coeff_df_meth_p$coeff > 0, "positive", "negative")

coeff_df_meth_p_cutoff <- coeff_df_meth_p[coeff_df_meth_p$coeff > 0.05 | coeff_df_meth_p$coeff < -0.05,]

# Find the maximum absolute coefficient value
max_abs_coeff_meth_p <- max(abs(c(min(coeff_df_meth_p_cutoff$coeff), max(coeff_df_meth_p_cutoff$coeff))))

# Create the plot with short_taxa on the y-axis, log_contrast on the x-axis, and fill based on sign
p_coda_glmnet_meth_p <- ggplot(coeff_df_meth_p_cutoff, aes(x = coeff, y = reorder(Species, coeff), fill = sign)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#F8766D","#00BFC4","#7CAE00","#C77CFF")) +
  scale_x_continuous(limits = c(-max_abs_coeff_meth_p, max_abs_coeff_meth_p)) +
  #geom_text(aes(label = round(coeff, 2)), hjust = -0.1, size = 3) +
  #theme(axis.text.y = element_text(hjust = 1)) +
  labs(x = "Log-contrast coefficient", y = "Taxa", title = "") +
  theme_classic() +
  theme(legend.position = "none")
p_coda_glmnet_meth_p

```
**CH4** -- Create predictions plot
```{r}
# Prepare the data
predictions_meth_p <- data.frame(Value = coda_glmnet_meth_p$predictions)
predictions_meth_p$Group <- ifelse(predictions_meth_p$Value > 0, 'Low CH4', 'High CH4')

# Find the maximum absolute value
max_abs_value_meth_p_pred <- max(abs(c(min(predictions_meth_p$Value[predictions_meth_p$Group == "High CH4"]), max(predictions_meth_p$Value[predictions_meth_p$Group == "Low CH4"]))))

# # Plot the data and set the y-axis limits to be symmetrical around zero
p_predictions_meth_p <- ggplot(predictions_meth_p, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(colour = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#F8766D","#00BFC4","#7CAE00","#C77CFF")) +
  scale_y_continuous(limits = c(-max_abs_value_meth_p_pred, max_abs_value_meth_p_pred)) +
  theme_minimal() +
  labs(title = " ",
       x = "",
       y = "Balance value") +
  theme_classic() +
  theme(legend.position = "none")
p_predictions_meth_p

```
**CH4** -- Combine 
```{r}
p_coda_glmnet_plots_meth_p <- egg::ggarrange(p_coda_glmnet_meth_p, p_predictions_meth_p, labels = c("", ""), 
                         ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_glmnet_plots_meth_p)
 
```
**CH4** -- Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/signature-plot_meth_p.png", plot = p_coda_glmnet_meth_p, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_meth_p.png", plot = p_coda_glmnet_plots_meth_p, width = 6, height = 4, dpi = 300) 

```
**H2** -- Plot selected taxa and their estimated regression coefficients
```{r}
# Create a new data frame to create the plot
coeff_df_hydro_p <- data.frame(
  taxa = coda_glmnet_hydro_p$taxa.name,
  # short_taxa = short_taxa_meth,
  coeff = coda_glmnet_hydro_p$`log-contrast coefficients`
)

####
#### simplify names
####

# # Assuming your data is in a data frame called 'data'
# coeff_df_hydro_p <- data.frame(coeff_df_hydro_p %>%
#   group_by(Species) %>%
#   summarise(coeff = sum(coeff)))

# Add a column indicating whether log_contrast is positive or negative
coeff_df_hydro_p$sign <- ifelse(coeff_df_hydro_p$coeff > 0, "positive", "negative")

coeff_df_hydro_p_cutoff <- coeff_df_hydro_p[coeff_df_hydro_p$coeff > 0.05 | coeff_df_hydro_p$coeff < -0.05,]

# Find the maximum absolute coefficient value
max_abs_coeff_hydro_p <- max(abs(c(min(coeff_df_hydro_p_cutoff$coeff), max(coeff_df_hydro_p_cutoff$coeff))))

# Create the plot with short_taxa on the y-axis, log_contrast on the x-axis, and fill based on sign
p_coda_glmnet_hydro_p <- ggplot(coeff_df_hydro_p_cutoff, aes(x = coeff, y = reorder(Species, coeff), fill = sign)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#F8766D","#00BFC4","#7CAE00","#C77CFF")) +
  scale_x_continuous(limits = c(-max_abs_coeff_hydro_p, max_abs_coeff_hydro_p)) +
  #geom_text(aes(label = round(coeff, 2)), hjust = -0.1, size = 3) +
  #theme(axis.text.y = element_text(hjust = 1)) +
  labs(x = "Log-contrast coefficient", y = "Taxa", title = "") +
  theme_classic() +
  theme(legend.position = "none")
p_coda_glmnet_hydro_p

```
Create predictions plot
```{r}
# Prepare the data
predictions_hydro_p <- data.frame(Value = coda_glmnet_hydro_p$predictions)
predictions_hydro_p$Group <- ifelse(predictions_hydro_p$Value > 0, 'Low H2', 'High H2')

# Find the maximum absolute value
max_abs_value_hydro_p_pred <- max(abs(c(min(predictions_hydro_p$Value[predictions_hydro_p$Group == "High H2"]), max(predictions_hydro_p$Value[predictions_hydro_p$Group == "Low H2"]))))

# # Plot the data and set the y-axis limits to be symmetrical around zero
p_predictions_hydro_p <- ggplot(predictions_hydro_p, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(colour = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#F8766D","#00BFC4","#7CAE00","#C77CFF")) +
  scale_y_continuous(limits = c(-max_abs_value_hydro_p_pred, max_abs_value_hydro_p_pred)) +
  theme_minimal() +
  labs(title = " ",
       x = "",
       y = "Balance value") +
  theme_classic() +
  theme(legend.position = "none")
p_predictions_hydro_p

```
**H2** -- Combine 
```{r}
p_coda_glmnet_plots_hydro_p <- egg::ggarrange(p_coda_glmnet_hydro_p, p_predictions_hydro_p, labels = c("", ""), 
                         ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_glmnet_plots_hydro_p)
 
```
**H2** -- Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/signature-plot_hydro_p.png", plot = p_coda_glmnet_hydro_p, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_hydro_p.png", plot = p_coda_glmnet_plots_hydro_p, width = 6, height = 4, dpi = 300) 

```
**Outliers** -- look at CH4 and H2 outliers against all of the data
Assign "group" with metadata subset
```{r}
hydro_meth_put <- map
hydro_meth_put <- hydro_meth_put[complete.cases(hydro_meth_put),]

```

```{r}
# Create new column 'Range' based on 'SampleID'
hydro_meth_put$Range <- ifelse(hydro_meth_put$SampleID %in% c("950N", "950N2", "962P", "962P2", "921C", "921C2", "735P", "735P2"),
                              "Extreme", # "989N", "737P"
                              "Normal")

```
Subset species data
```{r}
# transpose asv/species/genus table for coda_glmnet() -- use species for original names, species_c4m for simplified names
# x_species_o <- data.frame(t(species_c4m), check.names = T) # simplified names
x_species_o <- data.frame(t(species), check.names = T) # original names

# subset asv/species/genus data to keep subgroup data
x_species_o <- x_species_o[row.names(x_species_o) %in% hydro_meth_put$SampleID, ]

y_range_o <- as.factor(hydro_meth_put$Range)
y_range_o

```

```{r}
# identification of microbial signatures
coda_glmnet_o <- coda_glmnet(x=x_species_o, y=y_range_o, nfolds = 3, lambda = "lambda.min")
coda_glmnet_o$`apparent AUC`
coda_glmnet_o

```
Extract glmnet results and add to new data frame
```{r}
keep_elem <- c(1, 2, 3, 5, 6, 7) # keep elements containing taxa numbers, names, log and AUC values, etc
coda_list_o <- list(coda_glmnet_o[keep_elem])
coda_df_o <- do.call(cbind.data.frame, coda_list_o)
View(coda_df_o)

```
Save coda_glmnet results as .csv file
```{r}
write.csv(coda_df_o, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/glmnet-results_outliers.csv", row.names = FALSE)

```
**Outliers** -- Plot selected taxa and their estimated regression coefficients
```{r}
# Create a new data frame to create the plot
coeff_df_out_nor <- data.frame(
  taxa = coda_glmnet_o$taxa.name,
  # short_taxa = short_taxa_meth,
  coeff = coda_glmnet_o$`log-contrast coefficients`
)

####
#### simplify names
####

# Assuming your data is in a data frame called 'data'
coeff_df_out_nor <- data.frame(coeff_df_out_nor %>%
  group_by(Species) %>%
  summarise(coeff = sum(coeff)))

# Add a column indicating whether log_contrast is positive or negative
coeff_df_out_nor$sign <- ifelse(coeff_df_out_nor$coeff > 0, "positive", "negative")

coeff_df_out_nor_cutoff <- coeff_df_out_nor[coeff_df_out_nor$coeff > 0.05 | coeff_df_out_nor$coeff < -0.05,]

# Find the maximum absolute coefficient value
max_abs_coeff_out_nor <- max(abs(c(min(coeff_df_out_nor_cutoff$coeff), max(coeff_df_out_nor_cutoff$coeff))))

# Create the plot with short_taxa on the y-axis, log_contrast on the x-axis, and fill based on sign
p_coda_glmnet_out_nor <- ggplot(coeff_df_out_nor_cutoff, aes(x = coeff, y = reorder(Species, coeff), fill = sign)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#C77CFF","#7CAE00","#F8766D","#00BFC4")) +
  scale_x_continuous(limits = c(-max_abs_coeff_out_nor, max_abs_coeff_out_nor)) +
  #geom_text(aes(label = round(coeff, 2)), hjust = -0.1, size = 3) +
  #theme(axis.text.y = element_text(hjust = 1)) +
  labs(x = "Log-contrast coefficient", y = "Taxa", title = "") +
  theme_classic() +
  theme(legend.position = "none")
p_coda_glmnet_out_nor

```
**Outliers** -- Create predictions plot
```{r}
coda_glmnet_o

# Prepare the data
predictions_out_nor <- data.frame(Value = coda_glmnet_o$predictions)
predictions_out_nor$Group <- ifelse(predictions_out_nor$Value > 0, 'Normal', 'Outlier')

# Find the maximum absolute value
max_abs_value_out_nor_pred <- max(abs(c(min(predictions_out_nor$Value[predictions_out_nor$Group == "Outlier"]), max(predictions_out_nor$Value[predictions_out_nor$Group == "Normal"]))))

# # Plot the data and set the y-axis limits to be symmetrical around zero
p_predictions_out_nor <- ggplot(predictions_out_nor, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(colour = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#7CAE00","#C77CFF","#F8766D","#00BFC4")) +
  scale_y_continuous(limits = c(-max_abs_value_out_nor_pred, max_abs_value_out_nor_pred)) +
  theme_minimal() +
  labs(title = " ",
       x = "",
       y = "Balance value") +
  theme_classic() +
  theme(legend.position = "none")
p_predictions_out_nor

```
**Outliers** -- Combine 
```{r}
p_coda_glmnet_plots_out_nor <- egg::ggarrange(p_coda_glmnet_out_nor, p_predictions_out_nor, labels = c("", ""), 
                         ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_glmnet_plots_out_nor)

# COMBINE ALL 
library(cowplot)
p_coda_glmnet_plots_panel <- plot_grid(p_coda_glmnet_plots_cu_o,
                                       p_coda_glmnet_plots_pnp_o,
                                       p_coda_glmnet_plots_hydro_p,
                                       p_coda_glmnet_plots_meth_p,
                                       p_coda_glmnet_plots_out_nor,
                                       labels = c("A", "B", "C", "D", "E"),
                                       ncol = 1)
p_coda_glmnet_plots_panel

```
**Outliers** -- Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/signature-plot_out_nor.png", plot = p_coda_glmnet_out_nor, width = 6, height = 7, dpi = 300)

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_out_nor.png", plot = p_coda_glmnet_plots_out_nor, width = 6, height = 4, dpi = 300) 

ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/coda4microbiome/sig_pred-plots_panel.png", plot = p_coda_glmnet_plots_panel, width = 6, height = 10, dpi = 300) 

```
